
\chapter{Rotational dynamics}
\label{chapter:rotationaldynamics}
In this Chapter, we introduce the concepts of torque and moment of inertia. These will allow us to apply Newton's Second Law to rotating systems. 

\begin{learningObjectives}{
 \item Understand how to use vector quantities for describing the kinematics of rotations.
 \item Understand how to use torque to determine angular acceleration.
 \item Understand the moment of inertia and to calculate it.
 \item Understand conditions for static and dynamic equilibrium.
 }
\end{learningObjectives}

\begin{opening}
\begin{MCquestion}{A question}
\item a choice
\item another choice %correct
\end{MCquestion}
\end{opening}

\section{Rotational kinematic vectors}
Recall that we can describe the motion of a particle along a circle of radius $R$ by using its angular position, $\theta$, its angular velocity, $\omega$, and its angular acceleration, $\alpha$. The angular position can be defined as the angle made by the position vector of the particles, $\vec r$, and the $x$ axis of a coordinate system whose origin is the centre of the circle.  The angular velocity is then the rate of the change of the angular position, and the angular acceleration is the rate of change of the angular velocity:
\begin{align*}
\omega &= \frac{d}{dt}\theta \\
\alpha &= \frac{d}{dt}\omega
\end{align*}

\section{Torques}

\section{Moment of inertia}


\newpage
\section{Summary}

\begin{chapterSummary}{
\item Something that was learned
}
\end{chapterSummary}

\newpage
\begin{importantEquations}
This is an important equation
\begin{align*}
E = mc^2
\end{align*}

\end{importantEquations}


\newpage
\section{Thinking about the material}
\subsection{Reflect and research}

\begin{enumerate}
\item Something to research more.
\end{enumerate}
\subsection{To try at home}

\begin{tQuestion}Try doing this \end{tQuestion}

\subsection{To try in the lab}

\newpage
\section{Sample problems and solutions}
\subsection{Problems}
\begin{problemParts}{A question\label{Q:chaptertitle:q1}}
\item How close can he get to the hurdle before he has to jump?
\item What maximum height does he reach?
\end{problemParts}

\newpage
\subsection{Solutions}
\begin{solution}{\ref{Q:chaptertitle:q1}}
{
the solution
}
\end{solution}

